---
title: "JiahuanLi_A02_CodingBasics"
author: "Jiahuan Li"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r day1, echo=TRUE, results=T}
#1. 
c1 <- seq(100, from = 1, by = 4)
c1
## c1: name of the sequence. 

#2. 
mean <- mean(c1)
## the mean of ci is 49

median <- median(c1)
## the median of c1 is also 49.

#3. 

if (mean > median){
  print ("True")
} else {
  print ("False")
}
## a simple if command. The reply is false, indication the mean is not greater than the median.

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5.

name <- c("student 1", "student 2", "student 3", "student 4")
typeof (name)
## name is a character vector

score <- c("85", "78", "45", "47")
typeof (score)
## score is a numeric vector

pass <- c(TRUE, TRUE, FALSE, FALSE)
typeof (pass)
## pass is a logical vector

#7.
combine <- cbind(name, score, pass)
student_records <- as.data.frame(combine)
## the dataframe is named "student records"

#8.
## the dataframe already has the column names because it is made from vectors
## rename the columns
names(student_records) <- c("Name", "Score","Pass")
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: in a matrix, only similar data types can be stored whereas in a data frame like this, it can contain different data types like characters, integers, or other data frames. Basically Data Frame is advanced version of a Matrix where various types of data input types can be stored.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#10.
Judgement <- function (v) {
  ifelse (v >= 50, "TRUE", "FALSE")
}

#11.
Judgement(student_records$Score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: the `ifelse` command worked. That is because this function is a type of vector operation and checks the condition for every element of a vector. If using the `if` and `else` command, only the first element of a vector will be checked and a warning will be reported.
